High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-07-21 doc/asplos2013/interface.tex
Minor edits

\subsection{Processors}
\label{subsec:procs}
Listing~\ref{lst:procapi} shows the interface for {\tt Processor} objects which allow for
the creation of parallel computations.  Processor provide a way of naming 
every computational unit in the machine.  We describe how to find processor handles
in Section~\ref{subsec:machmodel}.  In the case of our current interface 
processors are either individual cores or discrete GPUs, but this can be easily modified 
to include new processor types (line 10).   Processors support a single {\tt spawn}
operation (lines 13-14) that will launch a new {\em task} on that processor.
Note that because the spawn operation
is invoked on a processor handle, it is possible to launch a task on any
processor from anywhere in the system.  
The spawn call takes an optional event that must trigger before the task can begin.  
The spawn call returns an event that will trigger when the task completes.

@@

Between deffered locks and barriers (described in Section~\ref{subsec:events}) clients 
have access to the same set of synchronization primitves that they traditionally have
in both threading and bulk-synchronous interfaces.  However, these operations can now 
be composed with other asynchronous operations which is not possible in any other interface.
of which we are aware.

@@

\subsection{Physical Regions}
\label{subsec:phyreg}
In distributed machines with discrete memories, data movement often consists of more
than a simple copy operation.  Many applications perform operations and transformations on
their data in conjunction with data movement for higher performance.  One common example is 
that applications accumulate reduction operations in a buffer and as part of a copy operation 
the reduction operations are applied to a destination buffer on the receiving side of the copy.  
In order to support these kinds of conjoined operations in an asynchronous environment, our interface 
must be aware of the structure of data.  To describe the structure of data our interface 
uses {\em physical regions.}

A physical region is an allocation of data in a single memory in the memory hierarchy.  Physical
regions are grouped into {\em classes} which are sets of physical regions that share the
same number and size of elements, but don't necessarily share the same data layout.  Our
interface supports sub-classing, but the details are omitted due to space constraints.  The
interface only supports data movement operations between physical regions that are of
the same class (or between super- and sub-classes).
% Say something about dynamically modifying the number of elements in a class

@@

Clients can also associate an operation as part of a physical region.  One example of
associating an operation with a physical region is shown on lines 13-14.  This method
will create a physical region called a {\em reduction instance} that will be assoicated 
with the given {\tt ReductionOp}.  The interface for a {\tt ReductionOp} is shown on 
lines 26-33.  A reduction operation must have an {\tt apply} method, and can optionally 
implement two additional methods, {\tt init} and {\tt fold} that support additional 
optimizations described in Section~\ref{subsec:reducimpl}.  The two template parameters 
describe the base element type of the region ({\tt LHS}:left-hand-side) and the type of 
the argument to the reduction ({\tt RHS}:right-hand-side).  The reduction operation has 
total control over the choice of data layout for the reduction instance (interface not shown).
Reduction instances can be copied to other reduction instances with the same operation
or to regular physical regions of the same class.

@@

\subsection{Machine Interface}
\label{subsec:machmodel}
The last component of the interface allows for introspection of the current machine.
Listing~\ref{lst:machineapi} shows the machine interface.
The {\tt Machine} object is a singleton object that serves two purposes.  First, it
is the mechanism for initializing the runtime; at the start of a program the client 
creates a machine object, instantiates a task table mapping task IDs to function pointers,
and invokes the {\tt run} method (line 30-31).
Second, the machine object provides an interface for the client to 
discover properties of the underlying machine (lines 35-43).  The machine object maintains
sets of all the unique {\tt Processor} and {\tt Memory} handles in the system.  Processors
were described in Section~\ref{subsec:procs}.  For every memory in the system there also 
exists a {\tt Memory} handle with a unique ID.  Different memories often, but not always, 
imply a different address space.  Examples of different memories are described in
Section~\ref{sec:impl}.

@@

The client can query these two relations by the {\tt get\_*\_affinity} calls
(lines 40-43).

>>>>

\subsection{Processors}
\label{subsec:procs}
Listing~\ref{lst:procapi} shows the interface for {\tt Processor} objects which allow for
the creation of parallel computations.  Processor provide a way of naming 
every computational unit in the machine.  We describe how to discover processor handles
in Section~\ref{subsec:machmodel}.  In the case of our current interface 
processors are either individual cores or discrete GPUs, but this is easily modified 
to include new processor types (line 10).   Processors support a single {\tt spawn}
operation (lines 13-14) that will launch a new {\em task} on that processor.
Note that because the spawn operation
is invoked on a processor handle, it is possible to launch a task on any
processor from anywhere in the system.  
The spawn call takes an optional event that must trigger before the task can begin.  
The spawn call returns an event that will trigger when the task completes.

@@

Between deffered locks and barriers (described in Section~\ref{subsec:events}) clients 
have access to the same set of synchronization primitves that they traditionally have
in both threading and bulk-synchronous interfaces.  However, these operations can now 
be composed with other asynchronous operations which is not possible in any other interface
of which we are aware.

@@

\subsection{Physical Regions}
\label{subsec:phyreg}
In distributed machines with discrete memories, data movement often consists of more
than a simple copy operation.  Many applications perform operations and transformations on
their data in conjunction with data movement for higher performance.  One common example is 
that applications accumulate reduction operations in a buffer and as part of a copy operation 
the reduction operations are applied to a destination buffer on the receiving side of the copy.  
To support these kinds of conjoined operations in an asynchronous environment, our interface 
must be aware of the structure of data so it can perform accompanying operations with 
data movement operations.  To describe the structure of data our interface uses {\em physical regions.}

A physical region is an allocation of data in a single memory in the memory hierarchy.  Physical
regions are grouped into {\em classes} which are sets of physical regions that share the
same number and size of elements, but don't necessarily share the same data layout.  Our
interface supports sub-classing, but the details are omitted due to space constraints.  The
interface only supports data movement operations between physical regions that are of
the same class (or between super- and sub-classes).  Note, clients can still pass arbitrary arrays
of bits via the processor {\tt spawn} call.
% Say something about dynamically modifying the number of elements in a class

@@

Clients can also associate an operation as part of a physical region.  One example of
associating an operation with a physical region is shown on lines 13-14.  This instance
of the {\tt create\_instance} method
will create a physical region called a {\em reduction instance} that will be assoicated 
with the given {\tt ReductionOp} specified by the ID.  The interface for a {\tt ReductionOp} is shown on 
lines 26-33.  A reduction operation must have an {\tt apply} method, and can optionally 
implement two additional methods, {\tt init} and {\tt fold} that support additional 
optimizations described in Section~\ref{subsec:reducimpl}.  The two template parameters 
describe the base element type of the region ({\tt LHS}:left-hand-side) and the type of 
the argument to the reduction ({\tt RHS}:right-hand-side).  The reduction operation has 
total control over the choice of data layout for the reduction instance (not shown).
Reduction instances can be copied to other reduction instances with the same operation
or to regular physical regions of the same class.

@@

\subsection{Machine}
\label{subsec:machmodel}
The last component of the interface allows for introspection of the current machine.
Listing~\ref{lst:machineapi} shows the machine interface.
The {\tt Machine} object is a singleton object that serves two purposes.  First, it
is the mechanism for initializing the runtime; at the start of a program the client 
creates a machine object, instantiates a task table mapping task IDs to function pointers,
and invokes the {\tt run} method (line 30-31).
Second, the machine object provides an interface for the client to 
discover properties of the underlying machine (lines 35-43).  The machine object maintains
sets of all the unique {\tt Processor} and {\tt Memory} handles in the system.  Processors
were described in Section~\ref{subsec:procs}.  For every memory in the system there also 
exists a {\tt Memory} handle with a unique ID.  Different memories often, but not always, 
imply a different address space.  Examples of different memories are described in
Section~\ref{sec:impl}.

@@

The client can query these two relations by the {\tt get\_*\_affinity} calls
(lines 40-43) to discover properties of the machine.

### 2012-07-21 doc/asplos2013/intro.tex
Minor edits

To combat this problem, parallel programming libraries have introduced
asynchronous constructs allowing programmers to overlap phases.  For
example, MPI has asynchronous {\em send} and {\em receive} operations
for sending messages in parallel with computation \cite{MPI}.  There
are two problems with this approach.  First, not all phases support
asynchronous constructs.  Second, the asynchronous operations do not
compose.  There is no way to express that one asynchronous computation
should begin as soon as another asynchronous communication operation
has completed.  The result is that at some point there must be
blocking operations to coordinate synchronous and asynchronous phases.
If blocking calls are not accurately placed in the program they can
lead to processor stalls that expose latency.  Correct placement of
blocking calls can be dependent on input data, algorithmic decisions
like blocking factors, and the underlying hardware.  To hide as much
latency as possible and to minimize processor stalls, 
%regardless of inputs, algorithmic choices, or underlying hardware,
all three major aspects of parallel programs (computation, data movement,
synchronization) should be asynchronous and composable.

@@

\item We describe a physical region system that provides structural information about
data layout.  The structural information provided by physical regions enables an extensible
interface for conjoining operations and transformations with asynchronous data movement for 
further lantecy hiding (Section~\ref{subsec:phyreg}).

>>>>

To combat this problem, parallel programming libraries have introduced
asynchronous constructs allowing programmers to overlap phases.  For
example, MPI has asynchronous {\em send} and {\em receive} operations
for sending messages in parallel with computation \cite{MPI}.  There
are two problems with this approach.  First, not all phases support
asynchronous constructs.  Second, the asynchronous operations do not
compose.  There is no way to express that one asynchronous computation
should begin as soon as another asynchronous communication operation
has completed.  The result is that at some point there must be
blocking operations to coordinate synchronous and asynchronous phases.
If blocking calls are not accurately placed in the program they can
lead to processor stalls that expose latency.  Correct placement of
blocking calls can be dependent on input data, algorithmic decisions, 
and the underlying hardware.  To hide as much
latency as possible and to minimize processor stalls, 
%regardless of inputs, algorithmic choices, or underlying hardware,
all three major aspects of parallel programs (computation, data movement,
synchronization) should be asynchronous and composable.

@@

\item We describe a physical region system that provides structural information about
data layout.  The structural information provided by physical regions enables an extensible
interface for conjoining operations on data (e.g. reductions) with asynchronous data movement for 
further lantecy hiding (Section~\ref{subsec:phyreg}).
